Semi-quantitative petrographic characterization of Bronze Age quartzite tools – an exploratory approach

### 2022-08-25 analysis/paper/paper.Rmd
Updated manuscript

A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, Mjæreum -@mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the cost of potentially amplifying noise [@mark_workman21 pp. 351-397; @schmid_etal22]. This effect can be seen in fig. \@ref(fig:nir-savitzky-transform), where one of the centered spectra from the dark group is compared to its derivative counterpart. The SG transformation amplifies a number of significant peaks in the spectra, as well as a shift in the instrumentation filter at 1800 nm. Out of the total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were left in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise leaves. Despite this there is still a considerable difference in light absorbance between the dark and light group, as can be seen by comparing fig. \@ref(fig:nir-spectra-summary) and fig. \@ref(fig:nir-savitzky-transform):A.

@@

\linebreak
 
Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig. \@ref(fig:nir-peak-summary)) [@hunt77 p. 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current spectra the 2 200 nm band is typically the more intense band.

@@

```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
nir.pc1var <- nir.pc1var
```

@@

The loading lines for the PCA can be seen in figure \@ref(fig:nir-pca-loadings). The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material \@ref(fig:nir-pca-fig). Darker material cluster towards the positive side of PC 1, with lighter towards the negative. The light group also separate into two general groups along PC 2. Material that feature less impurities and some translucence group towards the top left of the plot, with white/light grey samples grouping in the bottom left.

@@

## Raman
There are two distinguishing features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

A third feature, present in a small subset of the samples, appears as two peaks spaced closely together at ~ 695 cm^-1^ and 795 cm^-1^. Studies of clay minerals [@kloprogge17] have identified a number of band regions in the lower wavenumbers related to Si, Al and OH vibrations, including the 700 - 800 cm^-1^ band. Considering the petrogenesis of quartzite and the likelihood that clay minerals are present in the material it is possible that this is what generates the feature in the current dataset, although measurements on iron ore minerals have also generated a response in the 600 - 700 cm^-1^ band region [@ramanaidou15].

@@

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 ( %) related to the quartz and D - G peaks. PC 2 and PC 3, however, mainly describes the remaining noise left in the spectra after detrending and preprocessing the data. Interestingly, the loading for PC 3 does include the Al-O-Si peaks around the 700 cm^-1^ - 800 cm^-1^ band, which helps in separating out those samples where this feature is present. The measurements with mainly quartz peaks gravitates towards the negative side of PC 1, and those where the graphite peaks are dominant towards the opposite side. It is also clear that the darker material feature more pronounced graphite peaks, whereas the lighter and more colourless objects mainly feature quartz peaks. As in the case of the NIR data, however, there are exceptions that break this trend, as well as a “muddled” middle group with both well-pronounced graphite and quartz peaks. When PC 3 ( %) is added the samples with quartz and Al-O-Si peaks gravitate towards the bottom left, separating themselves from the "quarts" group.

@@

## XRF
Four samples had dimensions that were < 10 mm in size of their smallest dimension, and were therefore excluded from the analysis as there may be complications when comparing these to larger samples [@shackley11b pp.9-10]. The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (Bal) that represents the remaining unquantifiable elements in a sample. The elements represented in this unquantifiable group are all those with an atomic number lower than ^12^Mg (REF). Univariate analysis shows that ^14^Si makes up the main bulk of the composition of the material, with a bimodal distribution (fig. \@ref(fig:xrf-violin-plot)). The majority of the readings resulted in a ^14^Si content of less than 50%, and about ¼ of the readings (~ %) featured a ^14^Si content closer to 90%. The “low” ^14^Si group also feature a Bal between 40 – 70%.

The two first components of the PCA have an eigenvalue above 1 and the cumulative explained variance is ~ %. The loadings can be be seen in figure \@ref(fig:xrf-loadings). In the score plot of PC 1 and PC 2 two large clusters can be seen (fig. \@ref(fig:xrf-score-plot)). The smaller cluster on the left side in the score plot includes the samples with a ^14^Si content higher than 50%, whilst the larger cluster includes those with lower ^14^Si content. It does not seem to be ^14^Si that is responsible for this clustering, however, but rather ^26^Fe and ^40^Zr that are the main contributors. The “low ^14^Si” cluster shows a trend along PC 2 of quartzite with a darker hue grouping towards the bottom. This may indicate a possible relation to the ^26^Fe content of the material as ^26^Fe-bearing minerals such as hematite can contribute to the dark colour in quartzite [@howard05 p.711]. The brecciated quartz samples largely group together at the top of the “low ^14^Si” cluster, with a couple ending up closer towards the “high ^14^Si” cluster. The white (N9) material is fairly evenly distributed within the right cluster.

The group of measurements that are somewhat separated from the “low ^14^Si” cluster at the bottom are all but one from the rock painting site of *Finnforsberget*, which is located next to the Skellefte river. The *Finnforsberget* artefacts were sampled twice with object 491 having one measurement closer to the “low ^14^Si” group and the other the “high ^14^Si” group with ^13^Al estimated at 25%. This artefact was sampled twice on each side, resulting in similar results both times.

@@

```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-loadings-scree.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/009-xrf-pca-loadings-scree.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
```

@@

```{r xrf-pca, fig.cap= "xrf score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/009-xrf-pca.png", error = FALSE)
```

@@

# O2PLS
 \linebreak
 \linebreak

@@

```{r o2pls-plot, fig.cap= "o2pls score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/010-O2PLS.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/010-O2PLS.png", error = FALSE)
```

@@

The most easily recognizable megascopic feature of petrographic material is its color, which has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g. @forsberg85 pp.60–64; @kaeck09 pp.53–55]. While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite. Indeed, destructive techniques are typically necessary in order to document the various metamorphic features characteristic of quartzite [@howard05; @prieto_etal19]. Non-destructive application of the above spectroscopic techniques does, however, bring to light some interesting characteristics of the material. Although there are trends that align with the perceived color of the objects in the PCA there are also exceptions to this rule, as well as a “muddled” grey zone of material that does not conform to the light and dark extremes. It is also worth mentioning that the color classification does not take the heterogeneous nature of the material into consideration.

Although no distinct enough groupings were observed as to make classification possible, a number of structural and chemical characteristics were identified. Despite the Raman PCA lacking descriptive power, due to the few features observed in the spectra, the presence of the “graphite peaks” may present an important piece of information in the identification of the potential geological origin of the material. Interestingly, mentions of graphitic quartzite can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day. Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, and further studies would be necessary in order to infer resource management strategies from this evidence.

>>>>

A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region^[For comparison, @mjaerum12 has put forth a chronology for bifacial points in southwestern Norway], there have been attempts [@baudou92 p.99; @forsberg85 p.5; @forsberg89], suggesting a period of use that falls between ca. 3 800 - 2 000 BP [@kaeck09 p.60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite [@baudou78].

@@

The spectra were evaluated using using Savitzky-Golay (SG) filtering, a smoothing technique  widely applied in spectroscopy [@chryssikos_gates17 pp.78-81]. 2^nd^ derivative SG can amplify weaker peaks in the spectra, although at the cost of potentially amplifying noise [@mark_workman21 pp. 351-397; @schmid_etal22]. An example of this can be seen in fig. \@ref(fig:nir-savitzky-transform), where one of the centered spectra from the dark group is compared to its derivative counterpart. The SG transformation amplifies a number of significant peaks in the spectra, as well as a shift in the instrumentation filter at 1800 nm. Out of a total of `r length(spec.d.total)` dark spectra `r length(spec.d.filtered)` were kept in the dark group after filtering out those spectra that feature no discernible peaks or mainly noise. Despite this there is still a considerable difference in light absorbance between the dark and light group, as can be seen by comparing fig. \@ref(fig:nir-spectra-summary) and fig. \@ref(fig:nir-savitzky-transform):A.

@@

Based on the evaluation of the spectra a couple of significant peaks can be identified at 1 400 nm, 1 900 nm, 2 200 nm, and 2 350 nm. These features are all the results of overtones and vibration combinations of hydroxyl (OH) groups, with the two first peaks at 1 400 nm and 1 900 nm being indicative of molecular water (fig. \@ref(fig:nir-peak-summary)) [@hunt77 p. 508-510; @sciuto_etal19]. OH groups can produce a response in a number of different fields in the same material, with metal-OH combinations occuring in the 2 000 nm - 2 500 nm region. As a general rule, the 2 200 nm and 2 350 nm peak positions have been related to aluminium (Al) and magnesium (Mg) combinations [@hunt77 p.509; @clark_etal90 p.12664]. While there are exceptions, a more intense 2 200 nm band has typically been related to AlOH, and a more intense 2 350 nm band to MgOH. In the current spectra the 2 200 nm band is typically the more intense band.

@@

```{r nir-pca, fig.width=20, fig.height=20, fig.cap= "PCA score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/004-nir-pca.R"), echo = FALSE}
quietly(source("./analysis/paper/004-nir-pca.R", encoding = "UTF-8"))
nir.pc1var <- nir.pc1var
nir.pc2var <- nir.pc2var
nir.pc3var <- nir.pc3var
```

@@

The loading lines for the PCA can be seen in figure \@ref(fig:nir-pca-loadings). The first principal component (PC) shows a strong trend (`r nir.pc1var` %) that seems to largely correlate with the absorbance of the material (fig. \@ref(fig:nir-pca-fig)). This can be seen in the trend of darker material clustering towards the positive side of PC 1, with lighter towards the negative. A slight difference in the clustering of light coloured material can also be seen along PC 2 (`r nir.pc2var` %). Material with a white or light grey hue gravitates towards the negative side (bottom left), while material that showcases translucence, and leaning more towards a yellow-beige hue, cluster towards the positive (top left). This can be seen more clearly in figure XX, which shows a score plot of PC 1 and PC 2 with examples of the points added to illustrate the differences in appearance of the material. As can be seen there are a couple of exemptions to the aforementioned trends, such as the translucent samples 260 (middle right) or 390 (top middle), possibly due to the darker impurities in the material.

While the explained variance of PC 3 is low, (`r nir.pc3var` %), it still contains a number of the aforementioned features visible in the spectra (inverse to PC 1). Adding PC 3 to a score plot (fig. \@ref(fig:nir-pca-fig):B), however,  there seems to be little new information gained apart from a couple of samples, mainly from the light group, that break away from the remaining samples.

@@

## Raman
There are two distinguishing features present in the Raman data (fig. \@ref(fig:raman-spectra-summary) & fig. \@ref(fig:raman-peak-ident)), one which can be related to quartz and another that may indicate the presence of carbon (C) structures. The three initial bands at ~ 128 cm^-1^, ~ 200 cm^-1^ and ~ 460 cm^-1^, are some of the more intense quartz bands [@zhong_etal21], with the strongest peak occurring at 460 cm^-1^ [@gillet_etal90]. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the expected position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

Aside from the above more intense features a couple of weaker peaks can also be seen in a number of samples. Some of these could potentially be additional quartz bands [@zhong_etal21], such as a weak peak at 1160 cm^-1^ that is somewhat obscured by the D band. Two peaks can also be found in the 700 - 800 cm^-1^ band range, and while the 700 cm^-1^ band has been found to be one of the aforementioned quartz bands, studies of clay minerals [@kloprogge17], and iron ore minerals [@ramanaidou15], have also generated a response related to Si, Al and OH vibrations within the 700 - 800 cm^-1^ region.

@@

```{r raman-pca-var, fig.cap= "raman pca", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/007-raman-pca.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/007-raman-pca.R", encoding = "UTF-8")))
raman.pc1var <- raman.pc1var
raman.pc2var <- raman.pc2var
raman.pc3var <- raman.pc3var
```

The resulting PCA model (fig. \@ref(fig:raman-pca)) shows a clear trend along PC 1 (`r raman.pc1var` %) related to the quartz and D - G peaks. Samples where the D/G peaks dominate in terms of intensity are clustered towards the positive side of PC 1, with those that lack D/G peaks clustered to the opposite side. As can be seen in both the plot, and the spectra, the darker samples typically feature more intense D/G bands. PC 2 (`r raman.pc2var` %) captures similar features as PC 1, due to the main features being a presence or lack of D/G bands, which causes the "arch" visible in the plot. The "extremes" of PC 1 gravitate to the negative side of PC 2, with those that feature both distinct quartz and D/G bands clustering towards the positive side. PC 3 (`r raman.pc3var` %), however, captures no real features in the data and mainly describes the remaining noise left in the spectra after detrending and preprocessing the data.

@@

```{r xrf-si, fig.cap= "xrf violin plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/008-xrf-violin-plot.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/008-xrf-violin-plot.R", encoding = "UTF-8")))

#extract samples above 80% Si 
Si_90 <- Points.xrf %>% 
  filter(`Si` > 90)
#total number of samples in dataset

#calculate percentage of samples above 80% Si
#si_perc <- round(((length(unique(Si_80$sample_id))/length(unique(Points.xrf$sample_id)))*100), 0)
si_90 <- length(unique(Si_90$sample_id))
```

```{r xrf-pca-var, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/009-xrf-pca-2.R", encoding = "UTF-8")))

xrf_var <- xrf.pc1var + xrf.pc2var + xrf.pc3var + xrf.pc4var
```

## XRF
The XRF spectrometer is not capable of detecting all of the common elements, and the software therefore calculates a balance (BAL) that represents the remaining unquantifiable elements in a sample. The elements represented in this group are all those with an atomic number lower than ~12~Mg (REF).

The analysis shows that ~14~Si makes up the main main bulk of the material, with ~13~Al recorded up to ~ 5% for some samples (fig. \@ref(fig:xrf-violin-plot)). Interestingly, the majority of the readings resulted in a ~14~Si content of less than 50%, as well as a BAL between 40 – 70%, meaning that a significant portion of the material remains unknown. This could both be due to the limits of the instrumentation, as well as the limited sampling. Out of a total of `r length(tot_uniq_samples)` samples `r si_90` measured a ~14~Si content of > 90%.

~23~V, ~39~Y, and ~30~Zn and ~25~Mn recorded a high number missing values (below LOD), although in the case ~25~Mn the values recorded reached up to 0.3%. As the analysis was carried out with field methods in mind, the objects were not cleaned as to remove potential remains from physical handling (e.g. P, S). Taking these two factors into consideration the PCA was limited to the following elements:  ~12~Mg, ~13~Al, ~14~Si, ~19~K, ~20~Ca, ~22~Ti, ~26~Fe, ~38~Sr and ~40~Zr.

The initial PCA model of the full dataset produced 5 outliers, which significantly impacted the interpretation of the data (sample 183, 193, 259, 359 and 400). These samples were therefore excluded and the PCA run again, which resulted in 4 PCs with an eigenvalue above 1 (fig. \@ref(fig:xrf-pca-loadings-scree)), and a cumulative explained variance of `r xrf_var`%. K-means cluster analysis was then performed on the PCA data in order to help identify potential groups exhibiting similar chemofacies. Using the elbow graph method 3 clusters were deemed suitable for the analysis.

As can be seen in the score plot of PC 1-2 (fig. \@ref(fig:xrf-pca-a)), the three K-means clusters remain separated with next to no overlap. Along PC 1 cluster 3 is marked by a higher ~26~Fe and ~40~Zr content on average, as well as a ~14~Si content lower than 50%. It mainly features samples with a dark hue, with the few samples from the light group being grey in color and arguably somewhere in between what might be considered a "dark" or "light" hue. Cluster 2 is separated out along PC 2, and includes only samples with a ~14~Si content above 90%, while ~26~Fe and ~40~Zr is low or below the LOD. Cluster 1 includes the main bulk of the light group, as well as all of the *brecciated quartz* samples. As in the case of cluster 3, apart from 1 outlier, the remaining samples in cluster 1 have a ~14~Si content below 50%. Except for ~22~Ti, ~26~Fe and ~40~Zr, the cluster features comparable values to cluster 3. Where it stands out is in the large amount of samples with ~38~Sr readings. ~38~Sr is a lithophile element that can substitute for ~20~Ca in sulphates, feldspars, mica and clays, and has been the focus of study of carbonate diagenesis/paragenesis [@ahijado_etal05;@banner95].

@@

```{r xrf-pca-loadings-scree, fig.cap= "xrf scree plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/009-xrf-pca-2.R"), echo = FALSE}
knitr::include_graphics("../figures/009-xrf-pca-load-scree.png", error = FALSE)
```

@@

```{r xrf-pca-a, fig.cap= "xrf score plot", echo = FALSE}
#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/009-xrf-pca-a.png", error = FALSE)
```

@@

```{r xrf-pca-b, fig.cap= "xrf score plot", echo = FALSE}
#quietly(eval(parse("./analysis/paper/009-xrf-pca.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/009-xrf-pca-b.png", error = FALSE)
```

```{r xrf-kmeans-boxplot, fig.cap= "xrf boxplot by kmeans cluster", echo = FALSE}
#quietly(eval(parse("./analysis/paper/009-xrf-kmeans-boxplot.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/009-xrf-kmeans-boxplot.png", error = FALSE)
```

<!--# O2PLS
 \linebreak
 \linebreak

@@

```{r o2pls-plot, fig.cap= "o2pls score plot", cache=TRUE, cache.extra=tools::md5sum("./analysis/paper/010-O2PLS.R"), echo = FALSE}
quietly(eval(parse("./analysis/paper/010-O2PLS.R", encoding = "UTF-8")))
knitr::include_graphics("../figures/010-O2PLS.png", error = FALSE)
```
-->

@@

The most easily recognizable megascopic feature of petrographic material is its color, which, together with texture, has been the main attribute used for grouping quartzite in North Swedish archaeological research [e.g. @forsberg85 pp.60–64; @kaeck09 pp.53–55]. While it is true that some inference of elemental content can be made based on color this is significantly more difficult with heterogeneous material such as quartzite. Indeed, destructive techniques are typically necessary in order to document the various metamorphic features in quartzite [@howard05; @prieto_etal19]. Non-destructive application of the above spectroscopic techniques does, however, bring to light some interesting characteristics of the material. Although there are trends that align with the perceived color of the objects there are also frequent exceptions to this rule. It is also worth mentioning that the color classification does not take the heterogeneous nature of the material into consideration.

Although no distinct enough groupings were observed as to make classification possible, a number of structural and chemical characteristics were identified. Despite the lack of explanatory power in the Raman PCA the features observed in the spectra are important for future studies. The presence of the “graphite peaks” is important to understanding the diagenetic background of the material, and thus its potential origin. Interestingly, mentions of "graphitic quartzite" can be found in an early 20th century description of an excursion to Storfjället, Tärna parish, by the Geological Society [@geologiska25], but otherwise does not seem to be a frequently used term in the modern day. Naturally, identifying the geological origin does not necessitate a direct link between that point and the procurement of the material, but it is nonetheless a key point in understanding the material flow within the region.

Taking the sampling of the XRF data into consideration, some interesting patterns did emerge from the analysis. The first being the bimodal distribution of ~14~Si
